---
title: "rangemap vignette"
output:
  github_document:
    toc: yes
    toc_depth: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

### Package description

The **rangemap** R package presents various tools to create species range maps based on occurrence data, statistics, and distinct shapefiles. Other tools of this package can be used to analyze environmental characteristics of the species ranges and to create high quality figures of these maps.

<br>

### Installing the package

**rangemap** is in a GitHub repository and can be installed and/or loaded using the following code (make sure to have Internet connection). 

```{r, include=TRUE, message=FALSE}
# Installing and loading packages
if(!require(devtools)){
    install.packages("devtools")
}

if(!require(rangemap)){
    devtools::install_github("marlonecobos/rangemap")
}
library(rangemap)
```

<br>

### Using the package functions

#### Setting R up

The following code chunk installs (if needed) and loads the R packages that will be used to get some data for performing the example analyses with the **rangemap** package. 

```{r, include=TRUE, message=FALSE, warning=FALSE}
# pacakges from CRAN
pcakages <- c("rgbif", "maps", "maptools", "raster")
req_packages <- pcakages[!(pcakages %in% installed.packages()[, "Package"])]
if (length(req_packages) > 0) {
  install.packages(req_packages, dependencies = TRUE)
}
sapply(pcakages, require, character.only = TRUE)

# package from github
if(!require(kuenm)){
devtools::install_github("marlonecobos/kuenm")
}
library(kuenm)
```

The working directory will also be defined in this part.

```{r, eval=FALSE, include=TRUE}
# working directory
setwd("C:/Users/Marlon/Documents/R/Borrar") # YOUR/WORKING/DIRECTORY
```

<br>

#### Simple graphical exploration of your data.

The *rangemap_explore* function generates simple figures to visualize species occurrence data in the geographic space before using other functions of this package. The figure created with this function helps to identify countries involved in the species distribution. Other aspects of the species distribution can also be generally checked here; for instace, disjunt distributions, dimension of the species range, etc. 

The function's help can be consulted usign the following line of code:

```{r, eval=FALSE, include=TRUE}
help(rangemap_explore)
```

An example of the use of this function is written below.

```{r, include=TRUE, cache=TRUE}
# getting the data from GBIF
species <- name_lookup(query = "Dasypus kappleri",
                       rank="species", return = "data") # information about the species

occ_count(taxonKey = species$key[14], georeferenced = TRUE) # testing if keys return records

key <- species$key[14] # using species key that return information

occ <- occ_search(taxonKey = key, return = "data") # using the taxon key

# keeping only georeferenced records
occ_g <- occ[!is.na(occ$decimalLatitude) & !is.na(occ$decimalLongitude),
             c("name", "decimalLongitude", "decimalLatitude")]
```

A simple figure.

```{r, include=TRUE, cache=TRUE}
# simple figure of the species occurrence data
rangemap_explore(occurrences = occ_g)
```

Same figure with country codes.


```{r, include=TRUE, cache=TRUE}
# simple figure of the species occurrence data
rangemap_explore(occurrences = occ_g, show_countries = TRUE)
```

<br>

#### Species ranges from buffered occurrences

The *rangemap_buff* function generates a species range for a given species by buffering provided occurrences using a user-defined distance. An approach to the species extent of occurrence (using convex hulls) and the area of occupancy according to the IUCN criteria are also generated. Shape files (occurrences, species range, extent of occurrence, and area of occupancy) can be saved in the working directory if it is needed. 

The function's help can be consulted usign the following line of code:

```{r, eval=FALSE, include=TRUE}
help(rangemap_buff)
```

An example of the use of this function is written below.

```{r, include=TRUE, cache=TRUE}
# getting the data from GBIF
species <- name_lookup(query = "Peltophryne empusa",
                       rank="species", return = "data") # information about the species

occ_count(taxonKey = species$key[1], georeferenced = TRUE) # testing if keys return records

key <- species$key[1] # using species key that return information

occ <- occ_search(taxonKey = key, return = "data") # using the taxon key

# keeping only georeferenced records
occ_g <- occ[!is.na(occ$decimalLatitude) & !is.na(occ$decimalLongitude),
            c("name", "decimalLongitude", "decimalLatitude")]

# buffer distance
dist <- 100000
save <- FALSE # TRUE if you want to save the shapefiles in the working directory 
name <- "Test"

buff_range <- rangemap_buff(occurrences = occ_g, buffer_distance = dist,
                            save_shp = save, name = name)
```


The function *rangemap_fig* generates customizable figures of species range maps using the objects produced by other function of this package. Let's see how the generated range looks like.

```{r, include=TRUE, cache=TRUE}
# creating the species range figure
rangemap_fig(buff_range, zoom = 1.2)

#dev.off() # for returning to default par settings
```

For further details see the function help. 

```{r, eval=FALSE, include=TRUE}
help(rangemap_fig)
```

<br>

#### Species ranges from boundaries

The *rangemap_bound* function generates a species range polygon for a given species by considering all the polygons of administrative entities in which the species has been detected. An approach to the species extent of occurrence (using convex hulls) and the area of occupancy according to the IUCN criteria are also generated. Shape files can also be saved in the working directory if needed. 

The function's help can be consulted usign the following line of code:

```{r, eval=FALSE, include=TRUE}
help(rangemap_bound)
```

Examples of the use of this function with most of its variants are written below.

<br>

##### Using only occurrences

Following there is an example in wich administrative areas will be selected using only occurrences. The *rangemap_explore* function will be used for obtainig a first visualization of the species distributional range.

```{r, include=TRUE, cache=TRUE}
# getting the data from GBIF
species <- name_lookup(query = "Dasypus kappleri",
                       rank="species", return = "data") # information about the species

occ_count(taxonKey = species$key[14], georeferenced = TRUE) # testing if keys return records

key <- species$key[14] # using species key that return information

occ <- occ_search(taxonKey = key, return = "data") # using the taxon key

# keeping only georeferenced records
occ_g <- occ[!is.na(occ$decimalLatitude) & !is.na(occ$decimalLongitude),
            c("name", "decimalLongitude", "decimalLatitude")]

# checking which countries may be involved in the analysis
rangemap_explore(occurrences = occ_g)

level <- 0
dissolve <- FALSE
save <- FALSE # TRUE if you want to save the shapefiles in the working directory 
name <- "test1"
countries <- c("PER", "BRA", "COL", "VEN", "ECU", "GUF", "GUY", "SUR", "BOL")

bound_range <- rangemap_bound(occurrences = occ_g, country_code = countries, boundary_level = level, 
                              dissolve = dissolve, save_shp = save, name = name)
```

Figure of the generated range.

```{r, include=TRUE, cache=TRUE}
# creating the species range figure
rangemap_fig(bound_range)

#dev.off() # for returning to default par settings
```

<br>

##### Using only administrative area names

Following there is an example in wich administrative areas will be selected using only the names of the administrative entities known to be occupied by the species. This approach may be useful in circumstances where goegraphic coordinates or aqurate localitie descriptions do not exist.

```{r, include=TRUE, cache=TRUE}
# administrative areas invloved
data("country_codes") #list of country names and ISO codes
head(country_codes)
# View(country_codes) # if you want to see the complete list

level <- 0
adm <- c("Ecuador", "Peru", "Venezuela", "Colombia", "Brazil") # If we only know the countries in wich the species is
dissolve <- FALSE
save <- FALSE # TRUE if you want to save the shapefiles in the working directory 
name <- "test2"
countries <- c("PER", "BRA", "COL", "VEN", "ECU", "GUF", "GUY", "SUR", "BOL")

bound_range1 <- rangemap_bound(adm_areas = adm, country_code = countries, boundary_level = level,
                              dissolve = dissolve, save_shp = save, name = name)
```

Map of the generated range.

```{r, include=TRUE, cache=TRUE}
# creating the species range figure
rangemap_fig(bound_range1)

#dev.off() # for returning to default par settings
```

<br>

##### Using occurrences and administrative areas

An example of using both occurrences and administrative areas for creating species ranges with the function *rangemap_bound* is presented below. This option may be useful when these two types of infromation complement the knowledge of the species distribution.

```{r, include=TRUE, cache=TRUE}
# getting the data from GBIF
species <- name_lookup(query = "Dasypus kappleri",
                       rank="species", return = "data") # information about the species

occ_count(taxonKey = species$key[14], georeferenced = TRUE) # testing if keys return records

key <- species$key[14] # using species key that return information

occ <- occ_search(taxonKey = key, return = "data") # using the taxon key

# keeping only georeferenced records
occ_g <- occ[!is.na(occ$decimalLatitude) & !is.na(occ$decimalLongitude),
            c("name", "decimalLongitude", "decimalLatitude")]

level <- 0
adm <- "Ecuador" # Athough no record is on this country, we know it is in Ecuador
dissolve <- FALSE
save <- FALSE # TRUE if you want to save the shapefiles in the working directory 
name <- "test3"
countries <- c("PER", "BRA", "COL", "VEN", "ECU", "GUF", "GUY", "SUR", "BOL")

bound_range2 <- rangemap_bound(occurrences = occ_g, adm_areas = adm, country_code = countries,
                              boundary_level = level, dissolve = dissolve, save_shp = save, name = name)
```

Map of the species range.

```{r, include=TRUE, cache=TRUE}
# creating the species range figure
rangemap_fig(bound_range2)

#dev.off() # for returning to default par settings
```

<br>

#### Species ranges from hull polygons

The *rangemap_hull* function generates a species range polygon for a given species by creating convex or concave hull polygons based on occurrence data. An approach to the species extent of occurrence (using convex hulls) and the area of occupancy according to the IUCN criteria are also generated. Shape files can be saved if needed. 

The function's help can be consulted usign the following line of code:

```{r, eval=FALSE, include=TRUE, cache=TRUE}
help(rangemap_hull)
```

Examples of the use of this function with most of its variants are written below.

<br>

##### Convex hulls

With the example provided below, a species range will be constructed using convex hulls. This range will be split based on two distinct algorithms of clustering: hierarchical and k-means. Convex hull polygons are commonly used to represent species ranges, however in circunstances where biogeographic barriers for the species dispersal exist, concave hulls may be a better option.

```{r, include=TRUE, cache=TRUE}
# getting the data from GBIF
species <- name_lookup(query = "Dasypus kappleri",
rank="species", return = "data") # information about the species

occ_count(taxonKey = species$key[14], georeferenced = TRUE) # testing if keys return records

key <- species$key[14] # using species key that return information

occ <- occ_search(taxonKey = key, return = "data", limit = 2000) # using the taxon key

# keeping only georeferenced records
occ_g <- occ[!is.na(occ$decimalLatitude) & !is.na(occ$decimalLongitude),
             c("name", "decimalLongitude", "decimalLatitude")]

# unique polygon (non-disjunct distribution)
dist <- 100000
hull <- "convex" 
split <- FALSE
save <- FALSE # TRUE if you want to save the shapefiles in the working directory
name <- "test4"

hull_range <- rangemap_hull(occurrences = occ_g, hull_type = hull, buffer_distance = dist,
                            split = split, save_shp = save, name = name)

# disjunct distributions
## clustering occurrences with the hierarchical method
split <- TRUE
c_method <- "hierarchical"
split_d <- 1500000
name <- "test5"

hull_range1 <- rangemap_hull(occurrences = occ_g, hull_type = hull, buffer_distance = dist,
                            split = split, cluster_method = c_method, split_distance = split_d,
                            save_shp = save, name = name)

## clustering occurrences with the k-means method
c_method <- "k-means"
n_clus <- 3
name <- "test6"

hull_range2 <- rangemap_hull(occurrences = occ_g, hull_type = hull, buffer_distance = dist,
                            split = split, cluster_method = c_method, n_k_means = n_clus,
                            save_shp = save, name = name)
```

Now the figure of the species range.

```{r, include=TRUE, cache=TRUE}
# creating the species range figure
rangemap_fig(hull_range1) # try hull_range and hull_range2

#dev.off() # for returning to default par settings
```

<br>

##### Concave hulls 

With the following examples, the species range will be constructed using concave hulls. The species range will be calculated as an only area and as disjunt areas by clustering its occurrences using hierarchical and k-means algorithms.

```{r, include=TRUE, cache=TRUE}
# unique polygon (non-disjunct distribution)
dist <- 100000
hull <- "concave" 
split <- FALSE
save <- FALSE # TRUE if you want to save the shapefiles in the working directory
name <- "test7"

hull_range3 <- rangemap_hull(occurrences = occ_g, hull_type = hull, buffer_distance = dist,
                            split = split, save_shp = save, name = name)

# disjunct distributions
## clustering occurrences with the hierarchical method
split <- TRUE
c_method <- "hierarchical"
split_d <- 1500000
name <- "test8"

hull_range4 <- rangemap_hull(occurrences = occ_g, hull_type = hull, buffer_distance = dist,
                            split = split, cluster_method = c_method, split_distance = split_d,
                            save_shp = save, name = name)

## clustering occurrences with the k-means method
c_method <- "k-means"
n_clus <- 3
name <- "test9"

hull_range5 <- rangemap_hull(occurrences = occ_g, hull_type = hull, buffer_distance = dist,
                            split = split, cluster_method = c_method, n_k_means = n_clus,
                            save_shp = save, name = name)
```

Checking the figure.

```{r, include=TRUE, cache=TRUE}
# creating the species range figure
rangemap_fig(hull_range3) # try hull_range4 and hull_range5

#dev.off() # for returning to default par settings
```

<br>

#### Species ranges from ecological niche models

The *rangemap_enm* function.

The function's help can be consulted usign the following line of code:

```{r, eval=FALSE, include=TRUE}
help(rangemap_enm)
```

An example of the use of this function is written below.

```{r, include=TRUE, cache=TRUE}
# parameters
sp_mod <- raster::raster(list.files(system.file("extdata", package = "kuenm"),
                                    pattern = "sp_model.tif", full.names = TRUE))
sp_train <- read.csv(list.files(system.file("extdata", package = "kuenm"),
                                pattern = "sp_train.csv", full.names = TRUE))
occ_sp <- data.frame("A_americanum", sp_train)
thres <- 5
save <- FALSE # TRUE if you want to save the shapefiles in the working directory
name <- "test10"

enm_range <- rangemap_enm(occurrences = occ_sp, model = sp_mod, threshold_omission = thres,
                          save_shp = save, name = name)
```

Let's see how this range looks like.

```{r, include=TRUE, cache=TRUE}
# creating the species range figure
rangemap_fig(enm_range)

#dev.off() # for returning to default par settings
```

<br>

#### Species ranges using trend surface analyses

The *rangemap_tsa* function generates species range polygons for a given species using a trend surface analysis. An approach to the species extent of occurrence (using convex hulls) and the area of occupancy according to the IUCN criteria are also generated. Shape files can be saved in the working directory if it is needed. 

Trend surface analysis is a method based on low-order polynomials of spatial coordinates for estimating a regular grid of points from scattered observations. This method assumes that all cells not occupied by occurrences are absences; hence its use depends on the quality of data and the certainty of having or not a complete sampling of the regiong_of_interest.

The function's help can be consulted usign the following line of code:

```{r, eval=FALSE, include=TRUE}
help(rangemap_tsa)
```

An example of the use of this function is written below.

```{r, include=TRUE, cache=TRUE}
# getting the data from GBIF
species <- name_lookup(query = "Peltophryne fustiger",
                       rank="species", return = "data") # information about the species

occ_count(taxonKey = species$key[5], georeferenced = TRUE) # testing if keys return records

key <- species$key[5] # using species key that return information

occ <- occ_search(taxonKey = key, return = "data") # using the taxon key

# keeping only georeferenced records
occ_g <- occ[!is.na(occ$decimalLatitude) & !is.na(occ$decimalLongitude),
             c("name", "decimalLongitude", "decimalLatitude")]

# region of interest
WGS84 <- CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
w_map <- map(database = "world", regions = "Cuba", fill = TRUE, plot = FALSE) # map of the world
w_po <- sapply(strsplit(w_map$names, ":"), function(x) x[1]) # preparing data to create polygon
reg <- map2SpatialPolygons(w_map, IDs = w_po, proj4string = WGS84) # map to polygon

# other data
res <- 1
thr <- 0
save <- FALSE # TRUE if you want to save the shapefiles in the working directory
name <- "test11"

tsa_r <- rangemap_tsa(occurrences = occ_g, region_of_interest = reg, threshold = thr,
                    resolution = res, save_shp = save, name = name)
```

Let's take a look at the results.

```{r, include=TRUE, cache=TRUE}
# creating the species range figure
rangemap_fig(tsa_r)

#dev.off() # for returning to default par settings
```

<br>

#### Nice figures of species ranges

The *rangemap_fig* function can be used to plot not only the generated species ranges but also the extent of occurrence and the species records in the same map.

The function's help can be consulted usign the following line of code:

```{r, eval=FALSE, include=TRUE}
help(rangemap_fig)
```

Examples of the use of this function are written below.

##### Including extent of occurrence

```{r, include=TRUE, cache=TRUE}
# arguments for the species range figure
extent <- TRUE

# creating the species range figure
rangemap_fig(hull_range5, add_extent = extent)

#dev.off() # for returning to default par settings
```

<br>

##### Including occurrences

```{r, include=TRUE, cache=TRUE}
# arguments for the species range figure
occ <- TRUE

# creating the species range figure
rangemap_fig(hull_range5, add_occurrences = occ)

#dev.off() # for returning to default par settings
```

<br>

##### Including extent of occurrence and species records

```{r, include=TRUE, cache=TRUE}
# arguments for the species range figure
extent <- TRUE
occ <- TRUE

# creating the species range figure
rangemap_fig(hull_range5, add_extent = extent, add_occurrences = occ)

#dev.off() # for returning to default par settings
```

<br>

##### Including other components

```{r, include=TRUE, cache=TRUE}
# arguments for the species range figure
extent <- TRUE
occ <- TRUE
grid <- TRUE # grid
legend <- TRUE # leggend of objects included
scale <- TRUE # scale bar
north <- TRUE # north arrow


# creating the species range figure
rangemap_fig(hull_range5, add_extent = extent, add_occurrences = occ,
             grid = grid, legend = legend, scalebar = scale, 
             scalebar_length = 1000, northarrow = north)

#dev.off() # for returning to default par settings
```

<br>

##### Saving the figure

```{r, eval=FALSE, include=TRUE, cache=TRUE}
# arguments for the species range figure
extent <- TRUE
occ <- TRUE
grid <- TRUE # grid
legend <- TRUE # leggend of objects included
scale <- TRUE # scale bar
north <- TRUE # north arrow
save <-  TRUE


# creating the species range figure
range_map <- rangemap_fig(hull_range5, add_extent = extent, add_occurrences = occ,
                          grid = grid, legend = legend, scalebar = scale, 
                          northarrow = north, save_fig = save)

#dev.off() # for returning to default par settings
```

<br>

#### Species ranges and environmental factors

##### Species ranges on environmental factor maps

The *ranges_emaps* function represents one or more ranges of the same species on various maps of environmental factors (e.g. climatic variables) to detect implications of using one or other type of range regarding the environmental conditions in the area. 

The function's help can be consulted usign the following line of code:

```{r, eval=FALSE, include=TRUE}
help(ranges_emaps)
```

An example of the use of this function is written below.

```{r, eval=FALSE, include=TRUE, cache=TRUE}

```

<br>

##### Species ranges in the environmental space

The *ranges_espace* function generates a three dimensional comparison of a species' ranges created using distinct algortihms, to visualize implications of selecting one of them if environmental conditions are considered. 

The function's help can be consulted usign the following line of code:

```{r, eval=FALSE, include=TRUE}
help(ranges_espace)
```

An example of the use of this function is written below.

```{r, include=TRUE, cache=TRUE}
# getting the data from GBIF
species <- name_lookup(query = "Dasypus kappleri",
                       rank="species", return = "data") # information about the species

occ_count(taxonKey = species$key[14], georeferenced = TRUE) # testing if keys return records

key <- species$key[14] # using species key that return information

occ <- occ_search(taxonKey = key, return = "data") # using the taxon key

# keeping only georeferenced records
occ_g <- occ[!is.na(occ$decimalLatitude) & !is.na(occ$decimalLongitude),
             c("name", "decimalLongitude", "decimalLatitude")]

# range based on buffers
dist <- 500000

buff <- rangemap_buff(occurrences = occ_g, buffer_distance = dist)


# range based on concave hulls
dist1 <- 250000
hull1 <- "concave"

concave <- rangemap_hull(occurrences = occ_g, hull_type = hull1, buffer_distance = dist1)


# ranges comparison in environmental space
## list of ranges
ranges <- list(buff, concave)
names(ranges) <- c("buff", "concave")

## other data for environmental comparisson
vars <- getData("worldclim", var = "bio", res = 5)

## mask variables to region of interest
WGS84 <- CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
w_map <- map(database = "world", regions = c("Ecuador", "Peru", "Bolivia", "Colombia", "Venezuela",
                                             "Suriname", "Guyana", "French Guyana"),
             fill = TRUE, plot = FALSE) # map of the world
w_po <- sapply(strsplit(w_map$names, ":"), function(x) x[1]) # preparing data to create polygon
reg <- map2SpatialPolygons(w_map, IDs = w_po, proj4string = WGS84) # map to polygon

e <- extent(reg)
mask <- as(e, 'SpatialPolygons')

variables <- crop(vars, mask)

## comparison
occur <- TRUE
env_comp <- ranges_espace(ranges = ranges, add_occurrences = occur, variables = variables)

env_comp
# now play around, zoom in and rotate the figure
```

```{r, eval=FALSE, include=TRUE}
# saving this figure may be challenging, try using
# the following lines of code and check your download directory

op <- options() # save default options
options(viewer = NULL) # set viewer to web browser
name <- "ranges_space" # name for figure

# using web browser to save image
env_comp %>% htmlwidgets::onRender(
  paste("function(el, x)
          {var gd = document.getElementById(el.id);
          Plotly.downloadImage(gd, {format: 'svg', width: ", 1000, ", height: ",
        800, ", filename: ", paste("\'", name, "\'", sep = ""), "});
          }", sep = "")
)

Sys.sleep(2) # waiting for the execution
options(viewer = op$viewer) # restore viewer to old setting (e.g. RStudio)
```
