---
title: "rangemap short tutorial II"
author: "Marlon E. Cobos, Vijay Barve, Narayani Barve, and Alberto Jiménez-Valverde"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rangemap short tutorial II}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Species distribution ranges based on distinct algorithms

If you haven't seen the rangemap short tutorial I, take a look at it for more background on the **rangemap** R package. 

In this short example we will explore three of the functions of this package that allow users to generate species ranges based on buffer distances, convex and concave hulls, and administrative areas.

### Installing or loading the needed packages

```{r, message=FALSE, warning=FALSE}
# devtools would help us to install rangemap from GitHub
if(!require(devtools)){
  install.packages("devtools")
}

# installing rangemap if needed
if(!require(rangemap)){
  devtools::install_github("marlonecobos/rangemap")
  library(rangemap)
}

# rgibif would help us to get some species occurrence data in our example
if(!require(rgbif)){
  install.packages("rgbif")
  library(rgbif)
}
```

### Getting and exploring the species occurrence data

Lest's first get some data from the GBIF database. Of course you can use your own data instead.

```{r, cache=TRUE}
# getting the data from GBIF
species <- name_lookup(query = "Ictinia mississippiensis",
                       rank="species", return = "data") # information about the species

species$key # to see all the keys we can try

# key 4 was the one with georeferenced occurrences
occ_count(taxonKey = species$key[9], georeferenced = TRUE) 
key <- species$key[9] # using species key that return information

occ <- occ_search(taxonKey = key, return = "data") # using the taxon key

# keeping only species name, and coordinates
occ_g <- occ[!is.na(occ$decimalLatitude) & !is.na(occ$decimalLongitude),
            c("name", "decimalLongitude", "decimalLatitude")]

```

Now let's see how this species records are geographically distributed using the *rangemap_explore* function.

```{r, warning=FALSE, cache=TRUE}
# checking which countries may be involved in the analysis
rangemap_explore(occurrences = occ_g)
```

### Species distribution based on buffers

The function *rangemap_buff* generates a distributional range for a given species by buffering provided occurrences using a defined distance. See the function's help for more details.

```{r, cache=TRUE}
# Defining parameters
dist <- 300000 # buffer distance in meters

# Now we can create the species range based on buffers
buff_range <- rangemap_buff(occurrences = occ_g, buffer_distance = dist)
```

### Species range based on administrative areas

With the *rangemap_bound* function we can generate a distributional range for a given species by buffering provided occurrences using a defined distance. See the function's help for more details.

```{r, cache=TRUE}
# Defining parameters
level <- 1 # level of detail for administrative areas, 1 = states
dissolve <- FALSE # make this TRUE if you want polygons with no admininstrative limits
countries <- c("USA", "MEX") # ISO names of countries involved in the analysis

# Now we can create the species range based on administrative areas
bound_range <- rangemap_bound(occurrences = occ_g, country_code = countries,
                              boundary_level = level, dissolve = dissolve)
```

### Species distribution based on hull polygons

The function *rangemap_hull* generates a distributional range for a given species based on provided occurrences and two distinct type of hulls, convex and concave. See the function's help for more details.

```{r, cache=TRUE}
# Defining parameters
dist <- 200000
hull <- "concave"

# Now we can create the species range based on administrative areas
hull_range <- rangemap_hull(occurrences = occ_g, hull_type = hull, buffer_distance = dist)
```

### Seeing all our results

The function *rangemap_fig* will allow you to produce a nice figure for each of the species ranges. Lets put all the figures together and compare them. 


```{r, fig.show='hold', warning=FALSE}
rangemap_fig(buff_range, add_occurrences = TRUE, zoom = 2, northarrow = TRUE,
             range_color = "transparent")
rangemap_fig(buff_range, zoom = 2)
rangemap_fig(bound_range, zoom = 2)
rangemap_fig(hull_range, zoom = 2, legend = TRUE, legend_position = "bottomleft")
```

## Note

For saving the results of the analyses as shapefiles use the parameters save_shp and name in the functions used to generate the species ranges. For saving each figure in png or other formats see the *rangemap_fig* function's help.
